---
title: "identifying_source_genes"
output: 
  github_document: default
  html_document: default
---

-Run diamond blast of faa files of 18 Bxy strains in 9 PICs against nr database

 *./scripts/pangenome_analyses/source_genes/diamond_PIC.sh*
 
-Read in results determine appropiate pident and palign cutoff values to find hits to other bacterial taxa that are comparably to hits to Bxy strains and subset to hits to those above that threshold
 
-Combine with taxonomic table from gtdb to assign higher taxonomic ranks (Phyla,Order,Family) for hits to determine taxonomic distribution of high similarity hits
 
-Output with table with GeneID and taxonomic distribution to be used by figures_tables.Rmd

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir ='/stor/work/Ochman/alex/captive_ape_strain_genomes')
library(tidyverse)
```

```{r}
diamond_dir = 'results/pangenome/Bacteroides_xylanisolvens/diamond/'
all_files = list.files(file.path(diamond_dir))
(select_files = all_files[str_detect(all_files,'_blast.txt')])

# defining the appropriate pident and palign cutoff
infile = file.path(diamond_dir,select_files[1])
df = read.csv(infile,sep='\t',header=F)
colnames(df) = c('qseqid', 'sseqid', 'pident', 'length', 'qlen','mismatch', 'gapopen',
               'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore', 'salltitles')
df = df %>% mutate(Bxy_hit = if_else(grepl("xylan", salltitles),1,0))
Bxy_df = df %>% filter(Bxy_hit == 1) 
summary(Bxy_df$pident)
Bxy_df = Bxy_df %>% filter(pident>.98)
summary(Bxy_df$length/Bxy_df$qlen)

#apply 98% pident and 98% palign cutoff to hits
read_diamond <- function(infile){
  outfile = str_replace(infile,'_blast.txt','_blast98.txt')
  df = read.csv(infile,sep='\t',header=F)
  colnames(df) = c('qseqid', 'sseqid', 'pident', 'length', 'qlen','mismatch', 'gapopen',
               'qstart', 'qend', 'sstart', 'send', 'evalue', 'bitscore', 'salltitles')
  df = df %>% mutate(Bxy_hit = ifelse(str_contains('xylanisolvens',salltitles),1,0))
  
  df = df %>% filter(pident>98 & length/qlen > .98) 
  df = df %>% separate_rows(salltitles,sep='<>')
  df = df %>% mutate(taxonomy = stringr::str_extract(string = df$salltitles,
                              pattern = "(?<=\\[).*(?=\\])"),
                     taxonomy = sapply(str_split(taxonomy,'\\['), tail, 1),
                     taxonomy = str_replace(taxonomy,"Candidatus ",""),
                     taxonomy = str_replace(taxonomy,"uncultured ",""),
                     taxonomy = str_replace(taxonomy,"unclassified ",""),
                     palign = length/qlen,
                     Genus = sapply(str_split(taxonomy,'\\ '), head, 1),
                     Species = sapply(str_split(taxonomy,'\\ '), "[", 2),
                     Species = replace_na(Species,value='sp.'),
                     Genus_Species = paste0(Genus,'_',Species)) %>%
                     as.data.frame()
  df = df %>% group_by(qseqid,Genus_Species) %>% 
      slice_min(order_by = evalue, n=1,with_ties = F) 
  write_tsv(df,outfile)
  return(df)
}

#filter blast hits 
#read_diamond(file.path(diamond_dir,'P14.E4_blast.txt'))
#read_diamond(file.path(diamond_dir,'P21.6E_blast.txt'))
#read_diamond(file.path(diamond_dir,'P21.11A_blast.txt'))

#use bac120 taxonomy in GTDB to define higher taxonomic ranks for hits
bac = read_tsv('bin/bac120_taxonomy.tsv',col_names = F,col_types = cols()) %>%
  separate(X2,sep=';',into = c('Domain','Phylum','Class','Order','Family','Genus','Species'))
bac = bac %>% mutate(Phylum = str_replace(Phylum,'p__',''),
                     Order = str_replace(Order,'o__',''), 
                     Family = str_replace(Family,'f__',''), 
                     Genus = str_replace(Genus,'g__',''),
                     Species = str_replace(Species,'s__',''),
                     Species = str_replace(Species,' ','_')) %>%
                mutate(Phylum = recode(Phylum,
                                       Bacteroidota = 'Bacteroidetes',
                                       Firmicutes_A = 'Firmicutes',
                                       Actinobacteriota = 'Actinobacteria'))

#merge results for 3 representatives captive ape lineages
diamond = read_tsv(file.path(diamond_dir,'P14.E4_blast98.txt'),col_types = cols()) %>%
    add_row(read_tsv(file.path(diamond_dir,'P21.6E_blast98.txt'),col_types = cols())) %>%
    add_row(read_tsv(file.path(diamond_dir,'P21.11A_blast98.txt'),col_types = cols())) 

#merge with GTDB to get phylum, order and family designations
diamond_genus = diamond %>% 
  left_join(distinct(select(bac,Phylum,Order,Family,Genus)),by = 'Genus') %>%
  rename('GeneID' = 'qseqid') %>%
  select(GeneID,pident,palign,salltitles,Phylum,Order,Family,Genus,Species,Genus_Species) 

#assign taxonomy info for hist whose entry for Genus is not found in GTDB
diamond_notfound = diamond_genus %>% filter(is.na(Phylum))
sort(table(diamond_notfound$Genus))

reclassify <- function(df,taxanames,phylum_order){
  df_to_change = df %>% filter(Genus %in% taxanames)
  df = df %>% filter(!Genus %in% taxanames)
  df_to_change$Phylum <- phylum_order[[1]]
  df_to_change$Order <- phylum_order[[2]]
  df_to_change$Family <- NA
  df_to_change$Genus <- NA
  df_to_change$Species <- NA
  df=rbind(df,df_to_change)
  return(df)
  }

diamond_genus = reclassify(diamond_genus,
                           taxanames = c('Bacteroidia','Bacteroidetes'),
                           phylum_order = c('Bacteroidetes',NA))
diamond_genus = reclassify(diamond_genus,
                           taxanames = c('Flavobacteriales'),
                           phylum_order = c('Bacteroidetes','Flavobacteriales'))
diamond_genus = reclassify(diamond_genus,
                           taxanames = c('Bacteroidaceae','Bacteroidales','Muribaculaceae',
                                         'Paludibacteraceae','Porphyromonadaceae','Mediterranea',
                                         'Dysgonamonadaceae','Dysgonomonadaceae','Gallibacteroides',
                                         'Hallella','Avibacteroides','Limisoma','Rikenellaceae',
                                         'Caccoplasma'),
                           phylum_order = c('Bacteroidetes','Bacteroidales'))
diamond_genus = reclassify(diamond_genus,
                           taxanames = c('Firmicutes','Erysipelotrichia','Clostridia','Solobacterium',
                                         'Clostridiales','Lachnospiraceae','Oscillospiraceae',
                                         'Tyzzerella'),
                           phylum_order = c('Firmicutes',NA))
diamond_genus = reclassify(diamond_genus,
                           taxanames = c('Proteobacteria','Shigella','Dechloromonas',
                                         'Alphaproteobacteria','Caedimonadaceae'),
                           phylum_order = c('Proteobacteria',NA))
diamond_genus = reclassify(diamond_genus,
                           taxanames = c('bacterium','Bacteria','organism','Lentisphaeria'),
                           phylum_order = c(NA,NA))
diamond_genus = reclassify(diamond_genus,
                           taxanames = c('Microviridae','Podoviridae','Bacteriophage',
                                         'Siphoviridae','Myoviridae','CrAss-like','virus'),
                           phylum_order = c('Bacteriophage',NA))

diamond_notfound = diamond_genus %>% filter(is.na(Phylum))
sort(table(diamond_notfound$Genus))
table(diamond_genus$Phylum)

#for an individual GeneIDs, determine taxonomic distribution of hits to nr database
diamond_genus = diamond_genus  %>% group_by(GeneID) %>%
                     mutate(found_in_phage = if_else('Bacteriophage' %in% Phylum,1,0),
                     multiphyla =if_else(
                       length(setdiff(unique(Phylum),c('Bacteriophage',NA)))>1,1,0),
                     multiorder = if_else(length(setdiff(unique(Order),NA))>1|multiphyla==1,1,0),
                     multigenus = if_else(length(setdiff(unique(Genus),NA))>1|multiorder==1,1,0),
                     multispecies = if_else(
                       length(setdiff(Species,c('sp.',NA)))>1|multigenus==1,1,0),
                     singlespecies = if_else(length(setdiff(Species,c('sp.',NA)))==1,1,0))

diamond_genus = diamond_genus  %>% group_by(GeneID) %>%
                    mutate( 
                      #assign taxonomic categories
                      cat = ifelse(multiphyla==1,'multiphyla',
                            ifelse(multiorder==1,'multiorder',
                              ifelse(multigenus==1,'multigenus',
                                ifelse(multispecies==1,'multispecies',
                                  ifelse(singlespecies ==1,'singlespecies','unclassified'))))),
                      
                      #assign most conserved taxonomic rank
                      group = ifelse(multiphyla==1,'Bacteria',
                          ifelse(
                            multiorder==1,toString(setdiff(unique(Phylum),c(NA,'Bacteriophage'))),
                              ifelse(multigenus==1,toString(setdiff(unique(Order),NA)),
                                  ifelse(multispecies==1,toString(setdiff(unique(Genus),NA)),
                                      ifelse(singlespecies==1,
                                            paste(
                                                toString(setdiff(unique(Genus),NA)),
                                                toString(setdiff(unique(Species),c('sp.',NA))),
                                                sep='_'),
                                            'unclassified'))))),
                      #creates lists for individual taxonomic ranks
                      phylalist = toString(setdiff(unique(Phylum),NA)),
                      orderlist = toString(setdiff(unique(Order),NA)),
                      genuslist = toString(setdiff(unique(Genus),NA)),
                      specieslist = toString(setdiff(unique(Species),NA))
                      )
#some hits were unclassified ie hitting to bacteriophage or Bacteroides_sp. but no classified taxa
unclassified = diamond_genus %>% filter(group=='unclassified') 
unclassified = unclassified %>% mutate(group = if_else(!is.na(Genus),Genus_Species,Phylum))
classified = diamond_genus %>% filter(group!='unclassified')                                    
diamond_genus =  rbind(classified,unclassified) 

diamond_genus_short = diamond_genus %>%
  group_by(GeneID) %>%
    slice_max(order_by='pident',with_ties=F,n=1) %>% #get a single representative per GENEID
    select(-c(pident:singlespecies)) #remove columns specific to individual hits

write_tsv(diamond_genus_short,file.path(diamond_dir,'diamond_blast_taxonomy.txt'))

```

